---
title: "Test2"
author: "Paul Christmann"
date: '2022-07-12'
output: pdf_document
geometry: "left=1.5cm,right=1.5cm,top=1.5cm,bottom=1.5cm"
header-includes:
  - \usepackage[font={small,it}, labelfont={bf}]{caption}
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
#library(rmdformats)
## Global options
#options(max.print="120")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
#opts_knit$set(width=120)
#opts_knit$set(root.dir  = "~/")
```

### The expression of all TRAs associated with a tissue cannot be used to infer organ development

In this research, we attempt to draw conclusions about the developmental state of a tissue based on the expression of genes associated with it alone. Therefore, we analyzed the share of differentially expressed transcripts above a certain expression level over time, as shown in Fig. ???A. Furthermore, we observed trends within the median expression of all differentially expressed transcripts associated with a tissue (Fig. ???B). Since both metrics only showed in miniscule changes, we hypothesized that distinct, counteracting trends in expression exsisted within one tissue. Thus, k-means clustering was used to determine groups of TRAs with similar expression patterns. For each of these clusters, the median expression was plotted as shown in Fig. ???C. 

```{r}
library(tidyverse)
load("diff_genes_ann_0.01.RData")
load("embryo_df_tissues.RData")
```


```{r}
#Create dataframe with median expression levels for each transcript and week (combining the replicates)
embryo_df_tissues_median  = data.frame(Week4 = apply(embryo_df_tissues[1:3], 1, median), Week5 =  apply(embryo_df_tissues[4:6], 1, median), Week6 =  apply(embryo_df_tissues[7:9], 1, median), Week7=apply(embryo_df_tissues[10:12], 1, median), Week8 =  apply(embryo_df_tissues[13:15], 1, median), Week9 =  apply(embryo_df_tissues[16:18], 1, median))
embryo_df_tissues_median = cbind(embryo_df_tissues_median, embryo_df_tissues[,c(19,32)])
tissue = "Spleen"
threshold = c(6.6,6.8, 7.0, 7.2, 7.4, 7.6, 8.0)

#Select the data from a tissues
bool_contained = sapply(embryo_df_tissues_median$tissues, function(x) {grepl(tissue, toString(x))})
data_temp = embryo_df_tissues_median[bool_contained,]

#Determine the percentage of expressed genes
count = 0
a = c()
data_plot = data.frame(Week = c(sapply(c(4:9), function(x) {rep(x, length(threshold))})), Threshold = factor(c(rep(threshold, 6))), Percentage = NA)
  for(s in 4:9) {
    for(t in 1:length(threshold)) {
      count = count +1
      a  = c(a,(sum(data_temp[,str_glue("Week", s)] > threshold[t]))/(dim(data_temp)[1]))
    }
  }
data_plot$Percentage= a

  
#Plot the data
spleen_percentage_plot =  ggplot(data_plot, aes(x = Week, y= Percentage, group = Threshold))   + geom_line(aes(colour = Threshold)) + geom_point(aes(colour = Threshold, shape = Threshold)) + labs(title = "Share over threshold") + theme_classic() + ylim(0,1)

```


```{r}
#Plot of median expression level of all spleen-related genes over time

#Select genes only of target tissue
bool_contained = sapply(diff_genes_ann_0.01$tissues, function(x) {"Spleen" %in% x})

#Graph of mean gene expression over time
spleen_genes = diff_genes_ann_0.01[bool_contained,]
spleen_genes = spleen_genes[, 1:19]
spleen_genes_mean = data.frame(week4=rowMeans(spleen_genes[,1:3]),week5=rowMeans(spleen_genes[,4:6]), week6=rowMeans(spleen_genes[,7:9]),week7=rowMeans(spleen_genes[,10:12]), week8=rowMeans(spleen_genes[,13:15]), week9=rowMeans(spleen_genes[,16:18]))
spleen_genes_graph = apply(spleen_genes_mean, 2, median)
spleen_genes_graph= data.frame (Expression = spleen_genes_graph, Week = c(4:9))

spleen_genes_plot = ggplot(spleen_genes_graph, aes(x = Week, y= Expression))  + geom_line(color = "#0000ff") + geom_point(color = "#0000ff") + labs(title = paste("Median expression"))  + theme_classic() +  ylim(6.5,8.5)

```


```{r}
#Clustered plot of expression levels over time for the spleen

library(factoextra)
library(cluster)
library(stringr)
library(ggplot2)
library(ggsci)
library(grid)
library(gridExtra)

tissue = "Spleen"

#Functions for later use
expression_difference = function(x,y) {
  return ((x-y)/((x+y)*0.5))
}

avg_sil = function(k, data) {
  km = kmeans(data, centers = k, nstart = 25)
  ss = silhouette(km$cluster, dist_data)
  mean(ss[, 3])
}

create_plot_data = function(dataset) {
  data_plot = data.frame(Week = c(sapply(c(4:9), function(x) {rep(x, n_cluster)})), Cluster = as.character(c(rep(c(1:n_cluster), 6))), Expression = NA)
  count = 0
  for(s in 4:9) {
    for(t in 1:n_cluster) {
      count = count + 1
      values = dataset[dataset$Cluster == t, str_glue("Week", s)]
      data_plot[count, "Expression"] = median(values)
      data_plot[count, "Cluster"] = str_glue(data_plot[count, "Cluster"], " [", km$size[t], " genes]")
   }
  }
  return(data_plot)
}



#Select an modify the data used for kmeans
bool_contained = sapply(diff_genes_ann_0.01$tissues, function(x) {grepl(tissue, toString(x))})
data_temp = diff_genes_ann_0.01[bool_contained,]
data_abs = data.frame(Week4 = apply(data_temp[1:3], 1, median), Week5 =  apply(data_temp[4:6], 1, median), Week6 =  apply(data_temp[7:9], 1, median),
                        Week7=apply(data_temp[10:12], 1, median), Week8 =  apply(data_temp[13:15], 1, median), Week9 =  apply(data_temp[16:18], 1, median))
data = data.frame("5_4" = expression_difference(data_abs[,"Week5"], data_abs[,"Week4"]), 
                  "6_5" = expression_difference(data_abs[,"Week6"], data_abs[,"Week5"]),
                  "7_6" = expression_difference(data_abs[,"Week7"], data_abs[,"Week6"]),
                  "8_7" = expression_difference(data_abs[,"Week8"], data_abs[,"Week7"]),
                  "9_8" = expression_difference(data_abs[,"Week9"], data_abs[,"Week8"]))
row.names(data) = rownames(data_abs)
  
#Select optimum cluster number
data = scale(data)
dist_data = get_dist(data, method = "pearson")
n_cluster = which.max(sapply(c(2:10), function(k) {avg_sil(k, data)})) +1
  
#Create the data for our plot
km = kmeans(data, centers = n_cluster, nstart = 25)
data_plot_raw = cbind(data_abs, Cluster = km$cluster)
plot_data = create_plot_data(data_plot_raw)
  
#Create the Plot
spleen_clustered_plot =  ggplot(plot_data, aes(x = Week, y= Expression, group = Cluster))  + geom_line(aes(colour = Cluster)) +
          geom_point(aes(colour = Cluster, shape = Cluster)) + labs(title = "Clustered expression") +
          scale_color_npg() + theme_classic()  + ylim(6.5,8.5)
   
```

```{r, fig.width= 10, fig.height= 3, fig.cap= "A. For different expression thresholds the share of differentially expressed transcripts with higher expressions than the threshold is depicted. B. The median "}

library(ggplot2)
library(grid)
library(gridExtra)
library(cowplot)
library(ggpubr)
plot = arrangeGrob(grobs = list(spleen_percentage_plot, spleen_genes_plot, spleen_clustered_plot), nrow=1, widths = c(3.3,2.7, 4), heights = c(0.7))
plot_output = as_ggplot(plot) + draw_plot_label(label = c("A", "B", "C"), size = 15, x = c(0.03, 0.35, 0.63), y = c(0.13, 0.13, 0.13)) 
annotate_figure(plot_output, top = text_grob("Expression over time of Spleen-related differentially expressed transcripts", color = "black", face = "bold", size = 16))
```




```{r, eval = FALSE}
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(org.Hs.eg.db)
library(ggsci)
library(ggupset)
library(tidyverse)
library(gridExtra)
library(grid)
library(gt)
library(gtExtras)
library(ggplot2)
library(tidyverse)
library(cowplot)


#List the tissues
#tissues = c("Adipose", "Cells - Transformed fibroblasts", "Skin",  "Muscle - Skeletal", "Bladder", "Kidney - Cortex",  "Colon", "Esophagus", "Liver", "Minor Salivary Gland", "Pancreas", "Small Intestine - Terminal Ileum", "Spleen", "Stomach", "Breast - Mammary Tissue", "Cervix", "Fallopian Tube", "Ovary", "Uterus", "Vagina", "Prostate", "Testis", "Adrenal Gland", "Pituitary", "Thyroid",  "Artery", "Heart", "Lung", "Whole Blood", "Cells - EBV-transformed lymphocytes", "Brain - Spinal cord", "Brain", "Nerve - Tibial")
tissues = c("Adipose",  "Skin",  "Muscle - Skeletal", "Bladder",   "Colon", "Esophagus", "Liver", "Minor Salivary Gland", "Pancreas",  "Spleen", "Stomach", "Cervix", "Fallopian Tube", "Ovary", "Prostate", "Testis", "Adrenal Gland", "Pituitary", "Thyroid",  "Artery", "Heart", "Lung", "Whole Blood", "Brain", "Nerve - Tibial")
#tissues = c("Adipose", "Muscle - Skeletal", "Artery", "Heart")
#tissues = "Muscle - Skeletal"


for (tissue in tissues) {
  bool_contained = sapply(embryo_df_tissues$tissues, function(x) {grepl(tissue, toString(x))})
  embryo_df_tissue = embryo_df_tissues[bool_contained,]
  deGenes = as.character(embryo_df_tissue$ensembl.entrez[!duplicated(embryo_df_tissue$ensembl.gene)])
  geneUniverse = as.character(embryo_df_tissues$ensembl.entrez[!duplicated(embryo_df_tissues$ensembl.gene)])
  enr_GO_tissue1 = enrichGO(gene = deGenes, ont = "BP",  OrgDb ="org.Hs.eg.db", universe = geneUniverse, readable=TRUE, pvalueCutoff = 0.05)
  enr_GO_tissue = as.data.frame(enr_GO_tissue1)
  if (dim(enr_GO_tissue)[1] <10) {npath = dim(enr_GO_tissue)[1]} else {npath = 10}
  if(npath != 0) {
     tissue_table = enr_GO_tissue[1:npath,c(2,6)]
     colnames(tissue_table) = c("Function", "Adj_p_value")
     assign(str_glue(tissue,"_table"), gt(tissue_table)|>  tab_header(title= tissue) |> fmt_markdown(columns = c(Function)) |> gt_theme_538() |>
             fmt_scientific(columns = c(Adj_p_value), decimals = 0)) 
    } 
  else {
    assign(str_glue(tissue, "_table"), NA)
  }
  gtsave(get(str_glue(tissue, "_table")),str_glue(tissue, "_table.png"))
}
ff = sapply(tissues, function(x){str_glue(x, "_table.png")})

for (i in 1:25) {
  assign(str_glue("p_", i), ggdraw() + draw_image(ff[i]))
}
grid.arrange(p_1,p_2,p_3,p_4,p_6,p_7,p_8,p_9,p_10,p_11,p_12,p_13,p_14,p_15,p_16,p_17,p_18,p_19,p_20,p_21,p_22,p_23,p_24,p_25)

```


```{r, include = FALSE, eval = FALSE}
#Rearrange the selectet plots so that we can combine the table with a plot 

library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(org.Hs.eg.db)
library(ggsci)
library(ggupset)
library(tidyverse)
library(gridExtra)
library(grid)
library(gt)
library(gtExtras)
library(ggplot2)
library(cowplot)


#Spleen 
tissue = "Spleen"
selected = c(1,2,3,4,5)
#Overrepresentation Analysis
bool_contained = sapply(embryo_df_tissues$tissues, function(x) {grepl(tissue, toString(x))})
embryo_df_tissue = embryo_df_tissues[bool_contained,]
deGenes = as.character(embryo_df_tissue$ensembl.entrez[!duplicated(embryo_df_tissue$ensembl.gene)])
geneUniverse = as.character(embryo_df_tissues$ensembl.entrez[!duplicated(embryo_df_tissues$ensembl.gene)])
enr_GO_tissue1 = enrichGO(gene = deGenes, ont = "BP",  OrgDb ="org.Hs.eg.db", universe = geneUniverse, readable=TRUE, pvalueCutoff = 0.05)
enr_GO_tissue = as.data.frame(enr_GO_tissue1)
#Create the table of most significant functions
if (dim(enr_GO_tissue)[1] <10) {npath = dim(enr_GO_tissue)[1]} else {npath = 10}
tissue_table = enr_GO_tissue[1:npath,c(2,6)]
colnames(tissue_table) = c("Function", "Adj_p_value")
t = as.tibble(tissue_table)
gtsave((gt(t)|>  tab_header(title= tissue) |> fmt_markdown(columns = c(Function)) |> gt_theme_538() |>
fmt_scientific(columns = c(Adj_p_value), decimals = 0) |>  tab_style(style = list(cell_text(weight = "bold")), locations = cells_body(columns =Function, rows = selected))),str_glue(tissue, "_table.png"))
assign(str_glue(tissue, "_t"), (ggdraw() + draw_image(str_glue(tissue, "_table.png"))))
#Plot the chosen functions
data_plot = data.frame(Week = c(sapply(c(4:9), function(x) {rep(x, length(selected))})), Function =
                             as.character(c(rep(enr_GO_tissue$Description[selected], 6)), Expression=NA))
for (i in selected) {
  xx = strsplit(enr_GO_tissue$geneID[i], "/")
  bool_contained1 = sapply(diff_genes_ann_0.01$HGNC.symbol, function(x) {grepl(x, toString(xx))})
  bool_contained1[is.na(bool_contained1)] = FALSE
  data_temp = diff_genes_ann_0.01[bool_contained1,]
  data_abs = data.frame(Week4 = apply(data_temp[1:3], 1, median), Week5 = apply(data_temp[4:6], 1, median), Week6 =  
                              apply(data_temp[7:9], 1, median), Week7=apply(data_temp[10:12], 1, median), Week8 = 
                              apply(data_temp[13:15], 1, median), Week9 =  apply(data_temp[16:18], 1, median))
  expr_path = colMeans(data_abs[1:6,])
  data_plot[data_plot$Function == enr_GO_tissue$Description[i],"Expression"] = expr_path
}
assign(str_glue(tissue, "_plot"), ggplot(data_plot, aes(x = Week, y= Expression, group = Function))  + geom_line(aes(colour =Function), position = position_dodge(0)) + geom_point(aes(colour = Function, shape = Function), position = position_dodge(0)) + labs(title = tissue) +  theme_classic() + scale_color_npg()  + theme(legend.title = element_text(size = 9), legend.text = element_text(size = 7), aspect.ratio = 1, legend.position = "bottom", legend.spacing.y = unit(0.1, "cm"),legend.key.size = unit(0.4, "cm")) + guides(colour = guide_legend(nrow = length(selected), title.position = "top",byrow = TRUE), shape = guide_legend(nrow = length(selected), title.position = "top",byrow = TRUE)))



#Brain 
tissue = "Brain"
selected = c(1,5,7)
#Overrepresentation Analysis
bool_contained = sapply(embryo_df_tissues$tissues, function(x) {grepl(tissue, toString(x))})
embryo_df_tissue = embryo_df_tissues[bool_contained,]
deGenes = as.character(embryo_df_tissue$ensembl.entrez[!duplicated(embryo_df_tissue$ensembl.gene)])
geneUniverse = as.character(embryo_df_tissues$ensembl.entrez[!duplicated(embryo_df_tissues$ensembl.gene)])
enr_GO_tissue1 = enrichGO(gene = deGenes, ont = "BP",  OrgDb ="org.Hs.eg.db", universe = geneUniverse, readable=TRUE, pvalueCutoff = 0.05)
enr_GO_tissue = as.data.frame(enr_GO_tissue1)
#Create the table of most significant functions
if (dim(enr_GO_tissue)[1] <10) {npath = dim(enr_GO_tissue)[1]} else {npath = 10}
tissue_table = enr_GO_tissue[1:npath,c(2,6)]
colnames(tissue_table) = c("Function", "Adj_p_value")
assign(str_glue(tissue,"_table"), gt(tissue_table)|>  tab_header(title= tissue) |> fmt_markdown(columns = c(Function)) |> gt_theme_538() |>
fmt_scientific(columns = c(Adj_p_value), decimals = 0) |>  tab_style(style = list(cell_text(weight = "bold")), locations = cells_body(columns =Function, rows = selected)))
gtsave(get(str_glue(tissue, "_table")),str_glue(tissue, "_table.png"))
assign(str_glue(tissue, "_t"), (ggdraw() + draw_image(str_glue(tissue, "_table.png"))))
#Plot the chosen functions
data_plot = data.frame(Week = c(sapply(c(4:9), function(x) {rep(x, length(selected))})), Function =
                             as.character(c(rep(enr_GO_tissue$Description[selected], 6)), Expression=NA))
for (i in selected) {
  xx = strsplit(enr_GO_tissue$geneID[i], "/")
  bool_contained1 = sapply(diff_genes_ann_0.01$HGNC.symbol, function(x) {grepl(x, toString(xx))})
  bool_contained1[is.na(bool_contained1)] = FALSE
  data_temp = diff_genes_ann_0.01[bool_contained1,]
  data_abs = data.frame(Week4 = apply(data_temp[1:3], 1, median), Week5 = apply(data_temp[4:6], 1, median), Week6 =  
                              apply(data_temp[7:9], 1, median), Week7=apply(data_temp[10:12], 1, median), Week8 = 
                              apply(data_temp[13:15], 1, median), Week9 =  apply(data_temp[16:18], 1, median))
  expr_path = colMeans(data_abs[1:6,])
  data_plot[data_plot$Function == enr_GO_tissue$Description[i],"Expression"] = expr_path
}
assign(str_glue(tissue, "_plot"), ggplot(data_plot, aes(x = Week, y= Expression, group = Function))  + geom_line(aes(colour =Function), position = position_dodge(0)) + geom_point(aes(colour = Function, shape = Function), position = position_dodge(0)) + labs(title = tissue) +  theme_classic() + scale_color_npg()  + theme(legend.title = element_text(size = 9), legend.text = element_text(size = 7), aspect.ratio = 1, legend.position = "bottom", legend.spacing.y = unit(0.1, "cm"),legend.key.size = unit(0.4, "cm")) + guides(colour = guide_legend(nrow = length(selected), title.position = "top",byrow = TRUE), shape = guide_legend(nrow = length(selected), title.position = "top",byrow = TRUE)))

```

```{r, fig.height = 4, fig.width = 12,echo = FALSE, eval = FALSE}
lay = rbind(c(1,2,2,3,4,4), c(1,2,2,3,4,4), c(NA,2,2,NA,4,4))
grid.arrange(grobs = list(Spleen_t,Spleen_plot, Brain_t, Brain_plot), nrow = 1)
```


```{r}
#Get the spleen dataset with the clusters
expression_difference = function(x,y) {
  return ((x-y)/((x+y)*0.5))
}

avg_sil = function(k, data) {
  km = kmeans(data, centers = k, nstart = 25)
  ss = silhouette(km$cluster, dist_data)
  mean(ss[, 3])
}
bool_contained = sapply(diff_genes_ann_0.01$tissues, function(x) {grepl("Spleen", toString(x))})
  data_temp = diff_genes_ann_0.01[bool_contained,]
  data_abs = data.frame(Week4 = apply(data_temp[1:3], 1, median), Week5 =  apply(data_temp[4:6], 1, median), Week6 =  apply(data_temp[7:9], 1, median),
                        Week7=apply(data_temp[10:12], 1, median), Week8 =  apply(data_temp[13:15], 1, median), Week9 =  apply(data_temp[16:18], 1, median))
  data = data.frame("5_4" = expression_difference(data_abs[,"Week5"], data_abs[,"Week4"]), 
                  "6_5" = expression_difference(data_abs[,"Week6"], data_abs[,"Week5"]),
                  "7_6" = expression_difference(data_abs[,"Week7"], data_abs[,"Week6"]),
                  "8_7" = expression_difference(data_abs[,"Week8"], data_abs[,"Week7"]),
                  "9_8" = expression_difference(data_abs[,"Week9"], data_abs[,"Week8"]))
  row.names(data_abs) = data_temp$ensembl.transcript
  row.names(data) = rownames(data_abs)
  
  data = scale(data)
  dist_data = get_dist(data, method = "pearson")
  n_cluster = which.max(sapply(c(2:10), function(k) {avg_sil(k, data)})) +1
  
  km = kmeans(data, centers = n_cluster, nstart = 25)
  data_plot_raw = cbind(data_abs, Cluster = km$cluster)

  spleen_genes_clustered = data_plot_raw

```

```{r Spleen-expression-plot, fig.cap="A further look on the expression of transcripts in the up- and downregulated clusters shows that the upregulated transcripts are close to the minimum expression level between 6 and 7 in week 4 and showing expressions between 7 and 8.5 by week 9. In contrast, fig.height=3, fig.width=5}
library(ggplot2)
#Compare the initial expression values of both clusters
spleen_genes_clustered_1 = spleen_genes_clustered[spleen_genes_clustered$Cluster==1, ]
spleen_genes_clustered_2 = spleen_genes_clustered[spleen_genes_clustered$Cluster==2, ]
spleen_boxplot_data = data.frame(Expression = NA, Week = NA, Cluster = NA)
i = 0
for (genelist in list(spleen_genes_clustered_1$Week4, spleen_genes_clustered_1$Week9, spleen_genes_clustered_2$Week4,  spleen_genes_clustered_2$Week9)) {
  i = i+1
  spleen_df = data.frame(Expression = genelist, Week = factor(rep((if(i%%2 == 0) {9} else {4}),length(genelist))), Cluster = rep((if(i<3) {"Up"} else
    {"Down"}),length(genelist)))
  spleen_boxplot_data = rbind(spleen_boxplot_data, spleen_df)
}
spleen_boxplot_data = spleen_boxplot_data[-1,]
spleen_boxplot = ggplot(spleen_boxplot_data, aes(x = Cluster, y = Expression, fill = Week)) + geom_boxplot(position=position_dodge(0.8)) + theme_classic() + labs(title = "Expression of spleen-associated transcripts")
spleen_boxplot

```


```{r}
get_gene_info = function(transcript_ids){
  library(rentrez)
  library(XML)
  table = data.frame(Transcript = NA, Gene = NA, Protein = NA, Summary = NA)
  for(x in transcript_ids) {
    table = rbind(table, get_single_transcript(x))
  }
  table
}

get_single_transcript = function(single_transcript) {
   res = entrez_search(db = "gene", term = single_transcript)
   if(!length(res$ids)==0) {
     results = entrez_summary(db = "gene", id = res$ids)
      c(single_transcript, as.character(results[2]),  as.character(results[3]),  as.character(results[17]))
   }
   else{
        c(single_transcript, NA, NA, NA)
   }
}

spleen1_gene_description = get_gene_info(rownames(spleen_genes_clustered_1))
spleen2_gene_description = get_gene_info(rownames(spleen_genes_clustered_2))
```

```{r Downregulated-Genes-Spleen, fig.width=12}

# Table for downregulated genes

library(gt)
library(gtExtras)
library(ggplot2)
library(tidyverse)
library(svglite)
library(webshot)

Expression =  data.frame(c(1:length(spleen_genes_clustered_2[,1])))
Expression$Expression_over_time = sapply(c(1:length(spleen_genes_clustered_2[,1])), function(x) {list(as.numeric(c(as.numeric(spleen_genes_clustered_2[x,1]), as.numeric(spleen_genes_clustered_2[x,2]), as.numeric(spleen_genes_clustered_2[x,3]),
           as.numeric(spleen_genes_clustered_2[x,4]),as.numeric(spleen_genes_clustered_2[x,5]),as.numeric(spleen_genes_clustered_2[x,6]))))})

spleen2_gene_description_spl = cbind(spleen2_gene_description[-1,], Expression)
spleen2_gene_description_spl = spleen2_gene_description_spl[,-5]


spleen2_gene_description_sum = spleen2_gene_description_spl[c(2,3,7,8,9,15,20,22,23,25,27,29,31,34,35,37,38,39,40,41,43,45,46,50,54,56,57,58,59,64,65,67,69,72,73,75,77,78,83,113,114,116,121,122,143),]
functions_2 = c("Component of  the centromere", "Interphase chromosome condensation", "Chromosome condensation and stabilization", "Spindle microtubule organization", "Spindle dynamics", "Spindle checkpoint function", "Highly expressed during mitosis", "Central role in mitosis", "", "Nuclear movement in the cell cycle", "Stabilization of the mitotic spindle", "Regulator of mitosis", "Stem cell pluripotency", "Involved with cellular proliferation", "Mismatch repair", "Regulates G1/S transition", "Moves chromosomes", "Anaphase-promoting complex substrate", "Unwinds Holliday junction", "Induces apoptotic chromatin condensation", "Promotes cell growth", "Transcriptional activator in proliferation", "Associates with centromer-kinetochore complex", "Initiation of DNA replication", "Regulatory in the cell cycle", "Maintaining chromosome integrity in mitosis", "Stable spindle microtubule capture", "CpG methylation", "Initiation of DNA replication", "Subunit of M-phase promoting factor", "Involved in kinetochore function", "Cell cycle progression", "Cell cycle checkpoint regulator", "Spindle checkpoint function", "Spindle microtubule organisation", "Inducing cell cycle arrest", "Subunit of condensin complex","Cell division cycle protein", "DNA polymerase accessory", "Centromere-microtubule complex", "Mismatch repair","Anaphase-promoting complex substrate","Segregation of chromosomes", "Central role in mitosis", "Contributor to developmental programs")
spleen2_gene_description_sum$Summary = functions_2
table_spleen2 = gt(spleen2_gene_description_sum)|>  tab_header(title= "Spleen - Downregulated", subtitle = "List of differentially epxressed genes") |> fmt_markdown(columns = c(Transcript, Gene, Protein, Summary)) |> gt_theme_538() |> gt_plt_sparkline( column = Expression_over_time, type = "shaded")
table_spleen2
```





```{r}
library(gt)
spleen2_gene_description_sum
gt(spleen2_gene_description_sum)
packageVersion("gt")
mat = rbind(c(1,1,2), c(3,3,4))
gt(mat)|>  tab_header(title= "Spleen - Downregulated", subtitle = "List of differentially epxressed genes")
```

```{r}
islands_tbl <- 
  tibble(
    name = names(islands),
    size = islands
  ) %>%
  arrange(desc(size)) %>%
  slice(1:10)
islands_tbl
gt_tbl = gt(islands_tbl)
gt_tbl
```



