---
title: "Introduction to R Dinkelacker"
author: "David Jewanski"
date: "6 5 2022"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
#library(rmdformats)
## Global options
options(max.print="120")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=120)
opts_knit$set(root.dir  = "~/")
```

## Librarys

```{r, echo=TRUE}
library(affy)
library(vsn)
library(AnnotationDbi)
library(hgu133plus2hsenstcdf)
library(hgu133plus2hsenstprobe) 
library(limma)
library(pheatmap)
library(tidyverse)
library(ggplot2)
library(hexbin)

```

## Daten einlesen
```{r, echo=TRUE}
setwd("C:/Users/david/Documents/Data Analysis Lokal/Raw-Data/CEL_files")
data.GSE15744 <- ReadAffy()
data.GSE15744@cdfName <- "HGU133Plus2_Hs_ENST"
data.GSE15744
summary(data.GSE15744)
```


## Daten speichern
```{r, echo=TRUE}
setwd("C:/Users/david/Documents/Data Analysis Lokal/r_files/")
save.image(file="C:/Users/david/Documents/Data Analysis Lokal/r_files/data.GSE15744.rda")
```


## Check and save images


```{r, echo=TRUE}

image(data.GSE15744[,1], col=rainbow(100, start = 0, end = 0.75)[100:1]) 
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394496.pdf") #looks good 
image(data.GSE15744[,2], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394498.pdf") #looks good
image(data.GSE15744[,3], col=rainbow(100, start = 0, end = 0.75)[100:1]) 
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394500.pdf") #looks good
image(data.GSE15744[,4], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394501.pdf") #a bit brighter
image(data.GSE15744[,5], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394503.pdf") #looks good
image(data.GSE15744[,6], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394505.pdf") #looks good
image(data.GSE15744[,7], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394507.pdf") #looks good
image(data.GSE15744[,8], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394509.pdf") #looks good
image(data.GSE15744[,9], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394511.pdf") #looks good
image(data.GSE15744[,10], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394514.pdf") #looks good
image(data.GSE15744[,11], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394516.pdf") #looks good
image(data.GSE15744[,12], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394517.pdf") #looks good
image(data.GSE15744[,13], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394519.pdf") #looks good
image(data.GSE15744[,14], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394521.pdf") #looks good
image(data.GSE15744[,15], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394523.pdf") #looks good
image(data.GSE15744[,16], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394525.pdf") #looks good
image(data.GSE15744[,17], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394527.pdf") #very bright
image(data.GSE15744[,18], col=rainbow(100, start = 0, end = 0.75)[100:1])
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RAW/GSM394529.pdf") #looks good
```


## vsn Normalisierung und Plot

```{r, echo=TRUE}

GSE15744.vsnrma <- vsnrma(data.GSE15744)


```




## RAW - Data:   Remove file-endings (.CEL) from sample names in Rownames and Columnames
```{r, echo=TRUE}
filenames <- rownames(pData(data.GSE15744))
samples <- substr(filenames, 1,9)                                   # Only keeps 1st to 9th character in the string -> cuts .cel from name
rownames(pData(data.GSE15744)) = samples
rownames


orig <- colnames(exprs(data.GSE15744))
new <- substr(orig, 1, nchar(orig)-4)                               # Removes last 4 characters in the string
colnames(exprs(data.GSE15744)) = new
colnames(exprs(data.GSE15744))
```

# Normalized data: Rownames: Expression data: Remove file-endings (.CEL) from sample names
```{r, echo=TRUE}
filenames <- rownames(pData(GSE15744.vsnrma))
samples <- substr(filenames, 1,9)                                   # Only keeps 1st to 9th character in the string -> cuts .cel from name
rownames(pData(GSE15744.vsnrma)) = samples
rownames(pData(GSE15744.vsnrma))

orig <- colnames(exprs(GSE15744.vsnrma))
new <- substr(orig, 1, nchar(orig)-4)                               # Removes last 4 characters in the string
colnames(exprs(GSE15744.vsnrma)) = new
colnames(exprs(data.GSE15744))
```




#Save data
```{r, echo=TRUE}
save.image(file="C:/Users/david/Documents/Data Analysis Lokal/r_files/data.GSE15744_vsn.rda")

```





# Analysis of the Data

```{r, echo=TRUE}

GSE15744.vsnrma
#ERKLÄRUNG:
#Datentyp: Expressionset. Besteht aus assayData, protocolData, phenoData (restlichen sind leer) und Annotation
#assayData: sind alle Expressionsdaten der 18 Microarrays gespeichert. Aufrufen könnte man sie mit exprs(GSE15744.vsnrma), aber Achtung, sehr sehr sehr viele Daten
#phenoData, pData beschreiben Datensatz: 18 samples mit Namen und Metadaten
str(GSE15744.vsnrma) #gibt detailierte Infos über Datensatz, aber eher weniger hilfreich

dim(GSE15744.vsnrma)


```


# Plot: mean Sd plot
```{r, echo=TRUE}

meanSdPlot(GSE15744.vsnrma, xlab="Rank (mean)", ylab="Standard Deviation")
dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/vsn_mean_sd_plot.pdf")

```
# Plot: Boxplot
```{r, echo=TRUE}
## Before

boxplot(data.GSE15744, col = rainbow(15),
        cex.axis=0.5,
        las=2,                                            # Vertical x-axis text
        ylab="Relative expression",
        main="Raw Gene expression in human embryo (GSE15744, raw data)")

dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/Boxplot_GSE15744_Raw-Data.pdf")

## After

boxplot(exprs(GSE15744.vsnrma), col = rainbow(15),
        cex.axis=0.5,
        las=2,                                            # Vertical x-axis text
        ylab="Relative expression",
        main="Raw Gene expression in human embryo (GSE15744, normalized)")

dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/Boxplot_GSE15744_normalized.pdf")
```

# Plot: Histogramm
```{r, echo=TRUE}
# Before
hist(data.GSE15744,
     col=rainbow(10),
     ylim = c(0,0.7),
     ylab= "Density",
     xlab="Log Intensity",
     main = "Density function of log intensity in human embryo (raw data)")

dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/Histogramm_GSE15744_Raw-Data.pdf")

# After
eset <- exprs(GSE15744.vsnrma)

plot(density(eset[,1]), type="n", 
     xlab="log Intensity",
     ylim=c(0,0.7),
     ylab= "Density",
     main="Density function of log Intensity of human embryo (normalized data)")

for(i in 1:ncol(eset)){
  lines(density(eset[,i]), col=rainbow(10)[i])
}

dev.copy2pdf(file = "C:/Users/david/Documents/Data Analysis Lokal/Plots/Histogramm_GSE15744_vsnnormalized.pdf")
```


# Plot: RNA Degradation
```{r, echo=TRUE}
rnadeg.raw <- AffyRNAdeg(data.GSE15744)

plotAffyRNAdeg(rnadeg.raw, col=rainbow(10))
title(sub="Human embryo (GSE15744, Raw data)")

dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RNAdeg_GSE15744_raw.pdf")

# Not scaled
plotAffyRNAdeg(rnadeg.raw, col=rainbow(10), transform = "shift.only")
title(sub="Human embryo (GSE15744, Raw data)")

dev.copy2pdf(file="C:/Users/david/Documents/Data Analysis Lokal/Plots/RNAdeg_GSE15744_raw_unscaled.pdf")

# Samples look good: Lines don't cross
```


# Plot: Scatter Plot
```{r, echo=TRUE}
#jeweils ein Microarray wird mit dem nächsten Microarray verglichen.
for(i in 1:17){
  plot(eset[,c(i,i+1)], pch=".", cex=2)
  abline(0, 1, col="red")               # 45 degree dividing line
  
  title(main = paste("Scatterplot of probe", 
                     substr(colnames(GSE15744.vsnrma)[i], 1, nchar(colnames(GSE15744.vsnrma)[i])), "and", 
                     substr(colnames(GSE15744.vsnrma)[i+1], 1, nchar(colnames(GSE15744.vsnrma)[i+1])), 
                     sep=" ", collapse = NULL))
  
  file.name <- paste("C:/Users/david/Documents/Data Analysis Lokal/Plots/Scatterplots/Scatterplot_GSE15744.pdf", 
                     as.character(substr(colnames(GSE15744.vsnrma)[i], 1, nchar(colnames(GSE15744.vsnrma)[i]))), "_",
                     as.character(substr(colnames(GSE15744.vsnrma)[i+1], 1, nchar(colnames(GSE15744.vsnrma)[i+1]))),
                     ".pdf", sep="")
  
  dev.copy2pdf(file = file.name)
  dev.off()
}
```







