---
title: "Exploratory Data Analysis"
author: "Paul Christmann"
date: '2022-05-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load potentially useful libraries

library(Rcpp)
library(tidyverse)
library(affy)
library(vsn)
library(AnnotationDbi)
library(hgu133plus2hsenstprobe)
library(hgu133plus2hsenstcdf) 
library(limma)
library(pheatmap)
library(hexbin)
library(biomaRt)
```


```{r}
#Load Normalized Data (wihtout Affy control genes)

setwd("C:/Data Analysis")
load("data.vsnrma.rda")

```


```{r}
#Convert to dataframe and cut of Affymetrix Control genes

temp_embryo_df = as.data.frame(exprs(embryo.vsnrma))
embryo_df = temp_embryo_df[63:95721,]
```


```{r}
#Create a new data frames with the mean expression for each week

embryo_df_means = data.frame(week4=rowMeans(embryo_df[,1:3]),week5=rowMeans(embryo_df[,4:6]), week6=rowMeans(embryo_df[,7:9]),week7=rowMeans(embryo_df[,10:12]), week8=rowMeans(embryo_df[,13:15]), week9=rowMeans(embryo_df[,16:18]))

```

```{r}
#Calculate the standard deviation for each gene

embryo_expr_genes_sd = data.frame(sd = apply(X =embryo_df_means, 1, sd))


```


```{r}
#Work with the Standard Deviation Values: Histogram

ggplot(embryo_expr_genes_sd, aes(sd)) + geom_histogram()

```


```{r}
#Work with the Standard Deviation Values: Boxplot

ggplot(embryo_expr_genes_sd, aes(sd)) + geom_boxplot()

```


```{r}
#Work with the Standard Deviation Values: Histogram

ggplot(embryo_expr_genes_sd, aes(sd)) + geom_histogram()

```


```{r}
#Annotate the dataset
setwd("C:/Data Analysis/TRAdata")
ensembl = as.data.frame(read.csv(file="ensembl_103_human.txt", sep = ","))   # Load the ensembl annotation data

TranscriptID_orig = rownames(embryo_df)
ensembl = ensembl %>% distinct(Transcript.stable.ID, .keep_all = TRUE) # Remove duplicate transcripts

embryo_df["Transcript.stable.ID"] = gsub("[\\._].*$" ,"", rownames(embryo_df)) # Add column for mergign


embryo_df_ann = left_join(embryo_df, ensembl, by="Transcript.stable.ID")   # Combine the Datasets by theier Transcript ID
rownames(embryo_df_ann) <- embryo_df_ann$Transcript.stable.ID

setwd("C:/Data Analysis")
save(embryo_df_ann, file = "embryo_df_ann.RData")

```


```{r}
#Check the annotated dataset
summary(embryo_df_ann)
head(embryo_df_ann)

```

```{r}
#Import the TRAs

setwd("C:/Data Analysis/TRAdata")
tra_gtex = as.data.frame(read.csv(file = 'tra.2017.human.gtex.5x.table.tsv', sep = "\t"))

```


```{r}
#Analyse the TRAs

summary(tra_gtex)
head(tra_gtex)

```

```{r}
#Rename the Columns

colnames(embryo_df_ann) = c("Week4_1", "Week4_2", "Week4_3", "Week5_1", "Week5_2", "Week5_3", "Week6_1", "Week6_2", "Week6_3", "Week7_1", "Week7_2", "Week7_3", "Week8_1", "Week8_2", "Week8_3", "Week9_1", "Week9_2", "Week9_3", "Transcript.stable.ID", "Gene.stable.ID", "Chromosome.scaffold.name", "HGNC.symbol", "AFFY.HG.U133.Plus.2.probe")
#week = c(4, 4, 4, 5, 5, 5, 6, 6, 6, 7, 7, 7, 8, 8, 8, 9, 9, 9)
#replicate = c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3)


```


```{r}
#Differential gene expression: Limma Analysis

design = model.matrix(~ 0+factor(c(rep(1,1), rep(2,1), rep(3,1), rep(4,1), rep(5,1), rep(6,1))))
colnames(design) = c("week4", "week5", "week6", "week7", "week8", "week9")
design
fit = lmFit(embryo_df_ann[1:6], design)
contrast.matrix = makeContrasts(week5-week4, week6-week5, week7-week6, week8-week7, week9-week8, levels = design)
fit2 = contrasts.fit(fit, contrast.matrix)
fit2 = eBayes(fit2)
topTable(fit2)
```


```{r}
#Differential gene expression: Limma Analysis
Comparisons = data.frame(week=c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6), replicate = c(1,2,3,1,2,3,1,2,3,1,2,3,1,2,3,1,2,3))
Time = factor(Comparisons$week)
Probe = factor(Comparisons$replicate)
design = model.matrix(~Probe+Time)
fit = lmFit(embryo_df_ann[1:18], design)
fit = eBayes(fit, trend=TRUE, robust = TRUE)
results = decideTests(fit)
summary(results)
```


```{r}
#Limma results
plotMD(fit, coef= "Time5", status = results)

```







```{r}
#Differential gene expression: Limma Analysis with better comparisons
weeks = c("week4", "week5", "week6", "week7", "week8", "week9")
f = factor(c("week4", "week4", "week4", "week5", "week5", "week5", "week6", "week6", "week6", "week7", "week7", "week7", "week8", "week8", "week8", "week9", "week9", "week9"), levels =weeks)
design = model.matrix(~0+f)
colnames(design) =  weeks
fit_2 = lmFit(embryo_df_ann[1:18], design)
#contrast.matrix = makeContrasts(for(x in weeks) for (y in weeks) paste(x, "-", y), levels = design)
contrast.matrix = makeContrasts(week5-week4, week6-week5, week7-week6, week8-week7, week9-week8, levels = design)
fit_2b = contrasts.fit(fit_2, contrast.matrix)
fit_2b = eBayes(fit_2b, robust = TRUE)

```



```{r}
#Differential gene expression: Limma Analysis 2 Results

results = decideTests(fit_2b, adjust.method = "BH")
vennDiagram(results)
topTable(fit_2b, coef = 4, n=100)



```


```{r}
#Differential gene expression: Limma Analysis 2, week5 and 4
contrast_5.4 = makeContrasts(week6-week4, levels = design)
fit5.4 = contrasts.fit(fit_2, contrast_5.4)
fit5.4 = eBayes(fit5.4)
results5.4 = decideTests(fit5.4)
vennDiagram(results5.4)
summary(results5.4)
topTable(fit5.4, n=30)

```

```{r}
#Differential gene expression: Limma Analysis 2, between weeks
contrast_all = makeContrasts(week9-week8, week9-week7, week9-week6, week9-week5, week9-week4, week8-week7,week8-week6, week8-week5, week8-week4, week7-week6, week7-week5, week7-week4, week6-week5, week6-week4, week5-week4, levels = design)
fit_all = contrasts.fit(fit_2, contrast_all)
fit_all = eBayes(fit_all)
results_all = decideTests(fit_all)
summary(results_all)

```


{r}
#Differential gene expression: Time course
library(splines)
X <- ns(c(4:9, 4:9, 4:9), df=3)
Group <- factor(c("replicate1", "replicate2", "replicate3", "replicate1", "replicate2", "replicate3", "replicate1", "replicate2", "replicate3", "replicate1", "replicate2", "replicate3", "replicate1", "replicate2", "replicate3", "replicate1", "replicate2", "replicate3"), levels = c("replicate1", "replicate2", "replicate3"))
design_t <- model.matrix(~Group*X)
fit_t <- lmFit(embryo_df_ann[1:18], design_t)
fit_t <- eBayes(fit_t)




```{r}
#Differential gene expression: Limma Analysis analog Nazli

M = cbind(embryo_df_ann[,1]-embryo_df_ann[,4], 
          embryo_df_ann[,2]-embryo_df_ann[,5], 
          embryo_df_ann[,3]-embryo_df_ann[,6])

colnames(M) = paste(rep("Week5 vs Week4",3), as.character(1:3), "M", sep=".")

# Create and empty matrix, 
design=as.matrix(rep(1,3))
colnames(design)= "Week5-Week4"

#calculate the fit and thus p-values
fit_n= lmFit(M,design)
fit_n= eBayes(fit_n)

pvalue01= sum(p.adjust(fit_n$p.value,"BH")< 0.01)
pvalue01 #0

pvalue05= sum(p.adjust(fit_n$p.value,"BH")< 0.05)
pvalue05 #20

pvalue1= sum(p.adjust(fit_n$p.value,"BH")< 0.1)
pvalue1 #28

pvalue2= sum(p.adjust(fit_n$p.value,"BH")< 0.2)
pvalue2 #225

pvalue3= sum(p.adjust(fit_n$p.value,"BH")< 0.3)
pvalue3 #811

pvalue4= sum(p.adjust(fit_n$p.value,"BH")< 0.4)
pvalue4 #1719

pvalue5= sum(p.adjust(fit_n$p.value,"BH")< 0.5)
pvalue5 #5387



```

```{r}
library(GEOquery)
setwd("C:/Data Analysis/Daten")
data_paper = getGEO("GSE15744")
summary(data_paper)
```



```{r}
data_paper = data_paper[[1]]
summary(data_paper)
```

```{r}
fData(data_paper)
```


```{r}
#exprs(data_paper) = log2(exprs(data_paper))
boxplot(log2(exprs(data_paper)))
boxplot(embryo_df_ann[1:18])
data_paper_lg = data_paper
exprs(data_paper_lg) = log2(exprs(data_paper_lg))
boxplot(exprs(data_paper_lg))
```


```{r}
#Test limma with paper data
weeks = c("week4", "week5", "week6", "week7", "week8", "week9")
f = factor(c("week4", "week4", "week4", "week5", "week5", "week5", "week6", "week6", "week6", "week7", "week7", "week7", "week8", "week8", "week8", "week9", "week9", "week9"), levels =weeks)
design = model.matrix(~0+f)
colnames(design) =  weeks
fit_paper = lmFit(exprs(data_paper), design)
#contrast.matrix = makeContrasts(for(x in weeks) for (y in weeks) paste(x, "-", y), levels = design)
contrast.matrix_paper = makeContrasts(week5-week4, week6-week5, week7-week6, week8-week7, week9-week8, levels = design)
fit_paper_2 = contrasts.fit(fit_paper, contrast.matrix_paper)
fit_paper_2 = eBayes(fit_paper_2)

```



```{r}
#Test limma with paper data

results_paper = decideTests(fit_paper_2)
summary(results_paper)

```




embryo_




```{r}
#Try to clean the data

embryo_df_ann_cleaned = na.omit(embryo_df_ann)
summary(embryo_df_ann_cleaned)

```



```{r}
#Test cleaned data

fit_noNA = lmFit(embryo_df_ann_cleaned[1:18], design)
#contrast.matrix = makeContrasts(for(x in weeks) for (y in weeks) paste(x, "-", y), levels = design)
contrast.matrix_noNA = makeContrasts(week5-week4, week6-week5, week7-week6, week8-week7, week9-week8, levels = design)
fit_noNA2 = contrasts.fit(fit_noNA, contrast.matrix_noNA)
fit_noNA2 = eBayes(fit_noNA2)


```



```{r}
#Test limma with noNA data

results= decideTests(fit_noNA2)
summary(results)

```



```{r}
#Remove outliers
library(Routliers)

outliers_mad(embryo_df_ann[1])

```


```{r}
#Trying stuff

plotMDS(embryo_df_ann[1:18])
plotMDS(embryo_df_ann_cleaned[1:18])
plotSA(fit_2b)

duplicateCorrelation(embryo_df_ann[1:18], design) # Läuft nicht


```


```{r}
#Trying stuff
genas(fit_2b, coef = c(2,3), plot = TRUE)

```



```{r}
#Cheating for good table in the project proposal

contrast_9.4 = makeContrasts(week9-week4, levels = design)
fit9.4 = contrasts.fit(fit_2, contrast_9.4)
fit9.4 = eBayes(fit9.4)
results9.4 = decideTests(fit9.4)
vennDiagram(results9.4)
summary(results9.4)
diff_transcripts = topTable(fit9.4, n= 9022)

```


```{r}
#Cheating for good table in the project proposal
library(plyr)
tra_diff = tra_gtex[tra_gtex$ensembl.transcript %in% rownames(diff_transcripts), ]
#count(tra_diff$tissues)
#tra_diff$tissues
for (x in (1:length(tra_diff$tissues))) {
  z[x] = strsplit(tra_diff$tissues[x], " - ")
}

```

```{r}
tra_diff
```

```{r}
a = c()
for (x in (1:length(z))) {
 for (y in (1:lengths(z[x]))) {
  a = c(a, z[[x]][y])
 }
}
```


```{r}
library(plyr)
count(a)
```

```{r}
library(plyr)
length(a)
a
```

```{r}
library(plyr)


for (x in (1:length(a))) {
  f[x] = strsplit(a[x], "/")
}





b = c()
for (x in (1:length(f))) {
 #for (y in (1:lengths(f[x]))) {
  b = c(b, f[[x]][lengths(f[x])])
 #}
}

diff_tissue_count = count(b)

rownames(diff_tissue_count) = diff_tissue_count[,1]
#diff_tissue_count = diff_tissue_count[,-1]

```

```{r}
library(ggplot2)
 q <- ggplot(diff_tissue_count, aes(x = x, y = freq))  + geom_bar(stat = "identity") 
 q +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 q+  + scale_fill_gradientn(colors= topo.colors(6))
```



